---
output:
  pdf_document: default
  html_document: default
bibliography: money-happiness.bibtex
---

```{r prep, echo = FALSE, warning=FALSE, message=FALSE}
library(readxl)
library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(knitr)
```

```{r import-wdi, echo = FALSE,warning=FALSE}
wdi <- read_excel("../Original-Data/original-wdi.xlsx", 
                  sheet = "Data",
                  range = "A1:E15")
names(wdi) <- c("country.name", "country.code", "series.name", "series.code", "var")
wdi <- wdi %>%
  select(-series.code) %>%
  mutate(
    country = case_when(
      country.code =="CHN"  ~ 8,
      country.code == "IND" ~ 17,
      country.code == "IDN" ~ 18,
      country.code == "PAK" ~ 27,
      country.code == "RUS" ~ 31,
      country.code == "USA" ~ 40,
      country.code == "JOR" ~ 45)
  ) %>%
  select(-country.code) %>%
  mutate(series.name = 
           case_when(
             grepl("^GDP", series.name) ~ "inc",
             grepl("^General", series.name) ~ "exp"
             )
         ) %>%
spread(series.name, var)
```

```{r import-pew, echo = FALSE,warning=FALSE}
pew <- read_dta(file = "../Original-Data/importable-pew.dta") %>% 
  select(country, q2, q74) %>%
  rename(satis.score = q2, age= q74) %>%
  filter( satis.score != 11, satis.score != 12) %>%
  filter(age >= 21, age <= 70) %>%
  group_by(country) %>%
  mutate(country.n = n()) %>%
  filter(country.n >= 900) %>%
  select(-country.n) %>%
  mutate(age2 = age^2)
```

```{r processing, echo = FALSE,warning=FALSE}
df <- inner_join(wdi,pew, by = "country") %>%
  group_by(country.name) %>%
  summarise(cm_satis = mean(satis.score),
            inc = first(inc),
            exp = first(exp)) 

```

<!-- Begin title page LaTeX section -->

\begin{flushright}
Richard Ball\\
rball@haverford.edu\\
First version: 2014/06/17\\
This revision: 2014/06/17\\
\end{flushright}\bigskip
\begin{center}
\fbox{\begin{minipage}{0.9\textwidth}
\textit{This is a hypothetical paper written by two hypothetical individuals.
It was prepared for presentation at the BITSS Summer Institute, June 2-6, 2014,
to illustrate the TIER documentation protocol.}\par
\begin{center}
\bigskip{www.haverford.edu/TIER}
\end{center}
\end{minipage}}
\end{center}

\vspace{10mm}

\pagenumbering{gobble}
\begin{center}
\LARGE{Money, Happiness and the Midlife Crisis}
\end{center}

\vspace{40mm}

\begin{center}
\large{Ellsworth Dägg\par  
Lecturer in Economics and Horticultureé\par  
Wallywell University\bigskip

and\bigskip

Picard Janné\par  
Director of Research\par  
Joint Economic and Statistical Taskforce}\bigskip
\end{center}

\vspace{30mm}

\begin{center}
\large{June 3, 2014}
\end{center}
\newpage
\pagenumbering{arabic}
<!-- End title page LaTeX section -->

##I. Introduction##
This paper examines evidence from a multi-country survey of subjective well-being
(SWB) to investigate two questions. The first is whether the survey shows evidence of the
existence of a phenomenon that can be interpreted as a midlife crisis. The second is whether
there is any evidence of an association between a country's average SWB and each of two macro
indicators, GPD per capita and government consumption expenditure.
The policy relevance and intellectual significance of these questions has been verified by
introspection. Trust us.[^1]    

The regression analysis of the survey data reported in this paper suggests that, across the
lifespan, SWB follows the classic inverted U pattern, but turned upside down.[^2]  

The nature of the association we observe between country averages of SWB and GDP per
capita depends heavily on one influential case, namely the US. When the US is included in the
sample, one can argue for a positive association; if the US is excluded, there is weak evidence of
a negative association.  

The association between SWB and government consumption appears to be negative,
although the US is again an outlier. The evidence for the negative association becomes clearer
when the US is excluded.  

##II. Previous literature##
This paper contributes to large and still growing literature that uses surveys of SWB to
investigate a variety of socio-economic questions.  

The seminal work in this literature is @easterlin_1974, which focused on the association
between SWB and various notions of income and socio-economic status. By one interpretation,
Easterlin's results suggest that we should expect to find little association between average income
and average happiness in cross-sectional, country-level data.  

The pattern of SWB over the life cycle has been studied by @blanchflower_2007, who find evidence of a U-shaped trajectory. They apparently fail to recognize, however,
the deep resonance-of an anti-isomorphic nature-of the trajectory they observe with the
classic inverse U pattern.  

##III. Data##
The survey data used in this paper are taken from @pew_2002_data. This survey consisted of interviews of between about 500 and 2,500
individuals conducted in each of 44 countries in 2002.  

In 37 of the countries included in the Pew survey, the number of respondents was less
than 900. We chose to drop all observations from those 37 countries, which left us with a sample
of individuals from 7 countries: China, India, Indonesia, Jordan, Pakistan, the Russian Federation 
and the United States. Of the many variables based on respondents' answers to
social, political and economic questions, two were extracted for this paper. We gave these
variables the names SATIS and AGE.  

The variable SATIS is our measure of SWB. Values of this variable were elicited as
follows: Respondents were presented with a vertically ordered scale of integers, ranging from 10
at the top of the scale to 0 at the bottom, and the interviewer then asked:  

>Here is a ladder representing the "ladder of life." Let's suppose the top [a value of 10] of
the ladder represents the best possible life for you; and the bottom [a value of 0], the
worst possible life for you. On which step of the ladder do you feel you personally stand
at the present time? [@pew_2002_question, p. 6]    

The variable AGE represents the respondent's age in years at the time of her or his last
birthday. We also generated the variable AGE-SQUARED to use in the exploration of a possible
non-linear association between age and SWB.  

Two macro indicators were taken from @wdi_2014. GDP_PC shows the income
per capita of each country for 2002, expressed in current US dollars. GOV_CON shows
government consumption expenditure as a share of GDP, again for 2002.  

The variable CM_SATIS, representing country means of SATIS, was constructed for use
as a measure of SWB in country-level analyses.  

##IV. Descriptive statistics##
Table 1 shows the values of CM_SATIS, GDP_PC and GOV_CONS for each of the seven
countries in the sample.  

```{r table_1, echo=FALSE, results='asis', warning=FALSE}
# Mean Satisfaction, GPD per capita and Government Consumption by Country
# knitr::kable(df, caption = "Mean Satisfaction, GPD per capita and Government Consumption by Country")
# add these notes to table
# Notes:  
# a) GPD per capita in current (2002) US dollars  
# b) General government consumption expenditure, percent of GDP
df.tmp <- df
df.tmp[,-1] <-round(df.tmp[,-1],2)
names(df.tmp) <- c("Country", "Mean Satis", "GDP per capita", "Gov Cons")
kable(df.tmp, "latex", caption = "Mean Satisfaction, GPD per capita and Government Consumption by Country", booktabs = T) %>%
  #kable_styling(latex_options = "striped") %>%
  footnote(alphabet = c("GPD per capita in current (2002) US dollars", 
                        "General government consumption expenditure, percent of GDP"))
```

##V. Analysis and results##
To investigate whether the data contain evidence of the phenomenon of a midlife crisis,
we did an OLS regression of *SATIS* against *AGE* and *AGE_SQUARED*. The results of this
regression are shown in Column 1 of Table 2.  

We ran another OLS regression of *SATIS* against *AGE* and *AGE_SQUARED*, this time
including country fixed effects. The results of this regression are shown in Column 2 of 
Table 2.  

```{r table_2, echo = FALSE, warning = FALSE, results = 'asis', message=FALSE}
library(stargazer)
attach(pew)
fit <- lm(satis.score ~ age + age2)
stargazer(fit, type = 'latex', title = 'Regression Results')
detach(pew)
```

In both regressions, the coefficient on AGE is negative and the coefficient on and
*AGE_SQUARED* is positive, and in all four cases the p-values are below 0.0005. In both
regressions, therefore, the fitted value of *SATIS* follows an upside-down inverted-U 
pattern with respect to *AGE*, and reaches a unique interior minimum. We will call the age 
at which the minimum fitted value of *SATIS* occurs the "age of minimum SWB."  

If we use $\hat{\beta}_{AGE}$ to represent the estimated coefficient on *AGE* 
and $\hat{\beta}_{AGE-SQUARED}$ to represent the estimated coefficient on *AGE-SQUARED*, 
then the age of minimum SWB is equal to $-\frac{\hat{\beta}_{AGE}}{2\hat{\beta}_{AGE-SQUARED}}$. Using the estimated coefficients from 
our regressions we find:  

>Column 1 (no country fixed effects): Age at minimum SWB=47.76  
>Column 2 (with country fixed effects): Age at minimum SWB=53.90

Depending on which specification we use, we find that the age at minimum *SWB* is about
47 years and 9 months or 53 years and 11 months. We interpret this pattern, in which estimated
age at minimum SWB is within a few years of 50, as evidence in favor of a phenomenon that can
be called a midlife crisis.[^3]  

To examine the association between country means of SWB and GDP per capita, we
construct the scatterplot shown in Figure 1. The OLS line through those points would probably
be upward sloping, and maybe even statistically significant. However, such a result would be
influenced heavily by the observation for the US. It appears that if the US were dropped from the sample, 
the association might in fact be negative.  

To examine the association between country means of SWB and GDP per capita, we
construct the scatterplot shown in Figure 2. The OLS line through those points would probably
be downward sloping, and maybe even statistically significant. Again, however, the US is an
outlier. In this case, it appears that dropping the US would lead to a much cleaner negative
association.  

##VI. Conclusion##
The analysis presented in this paper provides evidence of an upside-down inverted U
relationship between SWB and age, with minimum SWB occurring within a few years of 50.
This result is consistent with the findings of Blanchflower and Oswald (2007).  

The country-level, cross-sectional associations between mean SWB and GDP per capita
and between mean SWB and government consumption expenditures depend heavily on whether
the US is included in the sample.

```{r fig_1, echo = FALSE}
fig1 <- ggplot(df, aes(x=inc, y=cm_satis))  +
  geom_text(label = df$country.name, vjust="top", hjust=-0.1) + 
  labs(title = "FIGURE 1",
       subtitle = "Scatterplot of GDP per capita and Mean Satisfaction") +
  expand_limits(x = 45000)
  
fig1 + geom_point() + labs(x = "GDP per capita (current [2002] $ US)", y = "Satisfaction")
```


```{r fig_2, echo = FALSE}
# Scatterplot of General Government Consumption and Mean Satisfaction
fig2 <- ggplot(df, aes(x=exp, y=cm_satis))  +
  geom_text(label = df$country.name, vjust="top", hjust=-0.1) +
  labs(title = "FIGURE 2",
       subtitle = "Scatterplot of General Government Consumption and Mean Satisfaction") +
  expand_limits(x = 25)
fig2 + geom_point() + labs(x = "Gov. cons., % of GDP", y = "Satisfaction")
```

##References## 


[^1]: This paragraph is a rip-off from Matthew Rabin, who made a similar quip during a seminar he gave at UC Berkeley
around 1990 or 1991 (about a new concept he was developing called fairness equilibrium).
[^2]: Commentators not well-versed in economics sometimes reveal their lack of sophistication by calling this pattern a U.
[^3]: A careful analysis would include also things like robustness tests and other forms of sensitivity analysis, as well as consideration of the magnitude or practical significance of the estimated coefficients. To maintain the realism of this paper as an example of economic research, such considerations are omitted.